---
title: "Population scaling"
vignette: >
  %\VignetteIndexEntry{Population scaling}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: false
    df-print: paged
    link-external-newwindow: true
    link-external-icon: true
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
execute: 
  eval: false
---

```{r}
#| label: setup

library(forestTIME)
library(dplyr)
library(rFIA)
library(ggplot2)
```

How can we use the interpolated data produced by `forestTIME` to get population-level (i.e. state-level) per-area estimates?

::: callout-warning

The interpolated values produced by `forestTIME` are *inferences* and not samples, so it may not be appropriate to use (modified) design-based estimators, as we do in this vignette, which treat the data, including interpolated values, as a probability sample.

:::

## Example data

We'll use RI as an example state because it is small. We'll compare our estimates with rFIA, adapting examples in the [FIA demystified vignette](https://doserlab.com/files/rfia/articles/fiademystified#without-sampling-errors) to our annualized data. We'll use the built-in example data that comes with `forestTIME`, but if you were to download your own and you wanted to use it with `rFIA` also, make sure to set `extract = "rFIA"` to extract all the tables needed for both `forestTIME` and `rFIA`.

```{r}
#| label: download-data
#| eval: false

fia_download(states = "RI", download_dir = "fia", extract = "rFIA")
```

::: callout-important
`rFIA` produces estimates of carbon from 33 -- 40.7 tons/acre using design-based estimators.
"Correct" estimates should be in this ballpark.

```{r}
#| label: rfia-biomass
rfia_RI <- readFIA(
  dir = system.file("exdata", package = "forestTIME"),
  states = "RI"
)

agc_rfia_annual <-
  biomass(
    rfia_RI,
    totals = TRUE,
    method = "annual",
    treeType = "live",
    landType = 'forest',
    component = "AG",
    areaDomain = COND_STATUS_CD == 1) |>
  mutate(method = "rFIA annual") |>
  select(method, YEAR, carbon_ton_acre = CARB_ACRE, carbon_total = CARB_TOTAL)

agc_rfia_ti <-
  biomass(
    rfia_RI,
    totals = TRUE,
    method = "TI",
    treeType = "live",
    landType = 'forest',
    component = "AG",
    areaDomain = COND_STATUS_CD == 1) |>
  mutate(method = "rFIA TI") |>
  select(method, YEAR, carbon_ton_acre = CARB_ACRE, carbon_total = CARB_TOTAL)

mean(agc_rfia_annual$carbon_ton_acre)
mean(agc_rfia_ti$carbon_ton_acre)
```

:::

We'll use the standard basic workflow to get estimated aboveground carbon for each tree in each year.  

```{r}
#| label: annualize-data

# Data prep
db <- fia_load(
  "RI",
  dir = system.file("exdata", package = "forestTIME")
)
data <- fia_tidy(db) #single tibble

# Expand to include all years between surveys, interpolate/extrapolate, and adjust for mortality
# Assign FIA design-based evaluation information to annualized data
# Estimate carbon
data_midpt <- data |>
  fia_annualize(use_mortyr = FALSE) |>
  fia_design(db) |>
  fia_allometry()
```

## Domain Indicator 

We'll add domain indicator columns as is done in the `rFIA` demystified vignette following equation 4.1, pg. 47 of Bechtold and Patterson(2005). These will be used to select which trees/plots/conditions fall within our population of interest. In this example we set the domain indicator for live trees per acre of forested land.

Reason:

> We build separate domain indicators for estimating tree totals and area totals, because we can specify different domains of interest for both.
> For example, if we used our tree domain (live trees on forest land) to estimate area, then we would not actually be estimating the full forested area in RI.
> Instead we would estimate the forested area ONLY where live trees are currently present.

So we can't just `filter(STATUSCD == 1 & COND_STATUSCD == 1)` to estimate carbon tons/acre.

```{r}
#| label: domain-indicators

data_midpt <-
  data_midpt |>
  mutate(
    aDI = if_else(COND_STATUS_CD == 1 & PLOT_STATUS_CD == 1, 1, 0), #forested land
    tDI = if_else(STATUSCD == 1 & PLOT_STATUS_CD == 1, 1, 0) * aDI #live trees on forested land
  )
```


## Without Sampling Error - Using Expansion Factors

Expansion factors are weights that scale tree-level measurements to population-level totals. They were widely used in the older periodic FIA sampling design because they made estimation simple. However, the expansion factors provided in FIADB (POP_STRATUM.EXPNS) do not support valid variance estimation, and in the annual panel system they become unstable because each panel combination produces different weights. They are still widely used so we will demonstrate how to use them with our annualized data set. See [Bechtold and Patterson (2005)](https://www.srs.fs.usda.gov/pubs/gtr/gtr_srs080/gtr_srs080.pdf) section 4.4 for more information. 

The FIA-supplied `EXPNS` values are no longer appropriate when using forestTIME because annualization changes the number of plots with "measurements" in each individual year. We therefore compute a new expansion factor (`EXPNS_ft`) in fia_design using FIA's post stratified evaluation information with our updated plot counts. 

We think these `EXPNS` values can be used much in the same way as the ones in the "raw" FIA database.

Using these expansion factors, we can follow the methods in the [FIA demystified vignette](https://doserlab.com/files/rfia/articles/fiademystified#without-sampling-errors).

```{r}
#| label: without-sample-error-estimators

tree_totals <- data_midpt |>
  group_by(plot_ID, YEAR) |>
  summarize(
    # purposefully omits adjustment factor `aAdj` because it is assumed to be 1
    carbPlot = sum(CARBON_AG_ft * TPA_UNADJ * EXPNS_ft * tDI / 2000, na.rm = TRUE), #tons/plot
  )

area_totals <- data_midpt |>
  group_by(plot_ID, YEAR) |>
  # Keep only one row for each condition in each plot and year
  distinct(CONDID, COND_STATUS_CD, CONDPROP_UNADJ, EXPNS_ft, aDI) |>
  summarize(
    # purposefully omits adjustment factor `aAdj` because it is assumed to be 1
    forArea = sum(CONDPROP_UNADJ * EXPNS_ft * aDI, na.rm = TRUE) #acres/plot
  )

agc_pop <- inner_join(tree_totals, area_totals) |>
  group_by(YEAR) |>
  summarize(
    CARB_AG_TOTAL = sum(carbPlot, na.rm = TRUE), # tons/plot
    AREA_TOTAL = sum(forArea, na.rm = TRUE) # acres/plot
  ) |>
  # the units work out to be tons(live carbon)/acre(forested land)
  mutate(method = "forestTIME", carbon_ton_acre = CARB_AG_TOTAL / AREA_TOTAL) |>
  select(
    method,
    YEAR,
    carbon_ton_acre,
    carbon_total = CARB_AG_TOTAL,
    AREA_TOTAL
  )
agc_pop
```

```{r}
#| label: plot-without-sample-error-results

all <- bind_rows(agc_rfia_annual, agc_rfia_ti, agc_pop)

ggplot(all, aes(x = YEAR, y = carbon_total, color = method)) +
  geom_line() +
  labs(y = "Total Carbon (tons)")

ggplot(all, aes(x = YEAR, y = carbon_ton_acre, color = method)) +
  geom_line() +
  labs(y = "Mean Carbon/Acre (tons/acre)")

```

We have some ballpark similar estimates to the ones `rFIA` produces.

## With Sampling Error

Note: Small strata merging has not yet been implemented in the forestTIME workflow so some strata may contain too few plots to compute reliable variance estimates. Small strata merging is currently being developed for forestTIME. Disregard if only interested in totals and ratio estimates.

If we want to compute sampling error using FIA's design-based variance estimator, we cannot use `EXPNS`. Instead, we need to use post-stratification where the landscape is divided into estimation units (for example, private land, public land, water) and then into smaller strata (for example, tree canopy cover) using remote-sensing data. Each pixel is assigned to a stratum, and the proportion of area in each stratum becomes its stratum weight. Phase 2 ground plots are then linked to their stratum (P2POINTCNT), and plot-level measurements are averaged within each stratum. The stratum means are then combined using the stratum weights first at the strata level and then at the estimation unit level to produced design-based population estimates. This stratified information changes from year to year as the land is resurveyed and this is reflected in the `EVALID`.

The `fia_design()` function attaches all of the necessary design information to the annualized data so that population estimates can be produced using FIA's post-stratified estimators. forestTIME attempts to match the data year to the same evaluation year however, some plots belong to non-matching evaluation years if a). no evaluation was done that year or b). the plot never belonged to that evaluation. 

```{r}
#| label: with-sample-error-estimators-plot-level

#  Estimate Tree attributes -- plot level
tree_pop <- data_midpt |>
  filter(EXPVOL == TRUE) |>
  distinct(ESTN_UNIT_CN, STRATUM_CN, CONDID, plot_ID, tree_ID, YEAR, .keep_all = TRUE) |>
  group_by(YEAR, ESTN_UNIT_CN, STRATUM_CN, plot_ID, EVALID) |>
  summarize(
    carbPlot = sum(CARBON_AG_ft * TPA_UNADJ * tDI / 2000, na.rm = TRUE),
    .groups = "drop"
  )

# Estimate Area attributes -- plot level
area_pop <- data_midpt |>
  filter(EXPCURR == TRUE) |>   # use EXPCURR flag
  distinct(ESTN_UNIT_CN, STRATUM_CN, CONDID, plot_ID, YEAR, .keep_all = TRUE) |>
  group_by(YEAR, P1POINTCNT, P1PNTCNT_EU, P2POINTCNT_ft, AREA_USED,
           ESTN_UNIT_CN, STRATUM_CN, plot_ID, EVALID) |>
  summarize(
    forArea = sum(CONDPROP_UNADJ * aDI, na.rm = TRUE),
    .groups = "drop"
  )

# Joining the two tables
carb_pop_plot <- left_join(tree_pop, area_pop)
```

```{r}
#| label: with-sample-error-estimators-strata-level

# Strata level estimates
carb_pop_strat <- carb_pop_plot |>
  group_by(YEAR, ESTN_UNIT_CN, STRATUM_CN, EVALID) |>
  summarize(aStrat = mean(forArea, na.rm = TRUE), # Area mean
            carbStrat = mean(carbPlot, na.rm = TRUE), # Carbon mean
            # We don't want a vector of these values, since they are repeated
            P2POINTCNT = first(P2POINTCNT_ft),
            AREA_USED = first(AREA_USED),
            # Strata weight, relative to estimation unit
            w = first(P1POINTCNT) / first(P1PNTCNT_EU),
            # Strata level variances
            aVar = (sum(forArea^2) - sum(P2POINTCNT * aStrat^2)) / 
                   (P2POINTCNT * (P2POINTCNT-1)),
            carbVar = (sum(carbPlot^2) - sum(P2POINTCNT * carbStrat^2)) / 
                      (P2POINTCNT * (P2POINTCNT-1)),
            # Strata level co-varainces
            carbCV = (sum(forArea*carbPlot) - sum(P2POINTCNT * aStrat *carbStrat)) / 
                     (P2POINTCNT * (P2POINTCNT-1)), 
            .groups = 'drop')

# Moving on to the estimation unit estimates
carb_pop_eu <- carb_pop_strat |>
  group_by(YEAR, ESTN_UNIT_CN, EVALID) |>
  summarize(aEst = sum(aStrat * w, na.rm = TRUE) * first(AREA_USED), # Mean Area
            carbEst = sum(carbStrat * w, na.rm = TRUE) * first(AREA_USED), # Mean carbon
             # Estimation unit variances
            aVar = (first(AREA_USED)^2 / sum(P2POINTCNT)) *
              (sum(P2POINTCNT*w*aVar) + sum((1-w)*(P2POINTCNT/sum(P2POINTCNT))*aVar)),
            carbVar = (first(AREA_USED)^2 / sum(P2POINTCNT)) *
              (sum(P2POINTCNT*w*carbVar) + sum((1-w)*(P2POINTCNT/sum(P2POINTCNT))*carbVar)),
            # Estimation unit covariances
            carbCV = (first(AREA_USED)^2 / sum(P2POINTCNT)) *
              (sum(P2POINTCNT*w*carbCV) + sum((1-w)*(P2POINTCNT/sum(P2POINTCNT))*carbCV)), 
            .groups = 'drop')
```

```{r}
#| label: with-sample-error-estimators-estn-level

# sum across Estimation Units for totals
carb_pop <- carb_pop_eu |>
  group_by(YEAR) |>
  summarize(AREA_TOTAL = sum(aEst, na.rm = TRUE),
            CARB_TOTAL = sum(carbEst, na.rm = TRUE),
            # Ratios
            CARB_ACRE = CARB_TOTAL / AREA_TOTAL,
            # Total samping errors
            AREA_TOTAL_SE = sqrt(sum(aVar, na.rm = TRUE)) / AREA_TOTAL * 100,
            CARB_TOTAL_SE = sqrt(sum(carbVar, na.rm = TRUE)) / CARB_TOTAL * 100,
            # Ratio variances
            carbAcreVar = (1 / AREA_TOTAL^2) * (sum(carbVar) + 
                          (CARB_ACRE^2) * sum(aVar) - 
                          2 * CARB_ACRE * sum(carbCV)),
            CARB_ACRE_SE = sqrt(sum(carbAcreVar, na.rm = TRUE)) / 
                                   CARB_ACRE * 100) |>
  mutate(method = "forestTIME") |>
  
  # Re ordering
  select(YEAR, 
         carbon_ton_acre = CARB_ACRE, 
         carbon_total = CARB_TOTAL, 
         AREA_TOTAL, 
         method)

mean(carb_pop$carbon_ton_acre)
```

```{r}
#| label: plot-with-sample-error-results

all <- bind_rows(agc_rfia_annual, agc_rfia_ti, agc_pop)

ggplot(all, aes(x = YEAR, y = carbon_total, color = method)) +
  geom_line() +
  labs(y = "Total Carbon (tons)")

ggplot(all, aes(x = YEAR, y = carbon_ton_acre, color = method)) +
  geom_line() +
  labs(y = "Mean Carbon/Acre (tons/acre)")

```
